***
#Basic analysis of weather events and their impact in public health and in the economy across the United States from 1950 to 2011.
***
Coursera Data Science Specialization  -  John Hopkins University

Course - Project: Reproducible Research - Peer Assessment 2  

Student: Francisco Jaramillo  

Date: Sunday, September 21, 2014

***

##Synopsis
This analysis is based in the data available at the US National Weather Service web site.   

The questions to answer are:   

1. Across the United States, which types of events (as indicated in the 'EVTYPE' variable) are most harmful with respect to population health?   

2.Across the United States, which types of events have the greatest economic consequences?   

The field EVTYPE has some historical non standard names. In order to aggregate properly the data, all the relevant values of EVTYPE had to be standardized as per the table number 2.1.1 Storm Data Event Table (page 6) from the document NWSI 10-1615 of AUGUST 17, 2007.   

The data is grouped by EVTYPE. Aggregates by EVTYPE are calculated for the fields: FATALITIES, INJURIES, PROPERTY DAMAGE, CROP DAMAGE and TOTAL ECONOMIC DAMAGE, and graphics are plotted.   

The results show Tornado is the most harmful weather event with respect to population health and is also the event that has had the greatest economic consequences across the US from 1950 to 2011   

***

##Preliminar
This preliminary section includes an auxiliary piece of code whose unique purpose is to format the text comments on the report. That code is original from Leek, Peng, Caffo on the Data Science JHU/Coursera program.
```{r, echo =TRUE}
#    Loading library and functions to format comments 
library(knitr)
library(stringr)
nicecomment <- function(..., skip_before=FALSE, skip_after=TRUE) {
    wrapped <- strwrap(str_c(..., sep = " "),
                        width = getOption("width") - 2)
     mes <- str_c("| ", wrapped, collapse = "\n")
     if(skip_before) mes <- paste0("\n", mes)
     if(skip_after) mes <- paste0(mes, "\n")
     message(mes)
 }

{nicecomment("The piece of code for the function 'nicecomment'",
             " was copied","from pretty_out - 'submitscript1.'; which is",    
             "a submission script for a previous assignment.")
nicecomment("I guess I have to formally cite here:")
nicecomment("Lines 15 to 22 were taken from the script: 'submitscript1.R'", 
            "belonging to Leek, Peng, Caffo on the Data Science JHU/Coursera program.",
            "It is used here as an auxiliary function, only to format the text of the comments.",
            "All the code relevant to the objectives of the assignment is original."
,skip_after=TRUE)
}
 
```



##Data Processing
The following Section loads the data and process for analysis.   

Loading is performed using download.file, bunzip2 and read.csv.    

This specific analysis has been structured based on the two main questions. So,in this analysis most of the main data processing will occur and the two following sections of this document. 
###Data Processing: Section: Loading the data


```{r, echo =TRUE}
#
#    Section: Loading the data
#         1. Creates a folder RepRes-proj2 in your working directory.
#         2. Download the data to RepRes-proj2 as a zip file from the course web site
#         3. Unzip the file
#         4. Load the data into a data frame named "data" on the R console 
#              by using the r function read.csv()


library(R.utils)
if(!file.exists("./RepData-proj2")){dir.create("./RepData-proj2")}
fileUrl <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileUrl,destfile="./RepData-proj2/data.bz2")
bunzip2("./RepData-proj2/data.bz2", destname=gsub("[.]bz2$", "", "./RepData-proj2/data.csv"), overwrite=TRUE, remove=FALSE)

data <- read.csv("./RepData-proj2/data.csv",header = TRUE, sep = ",", quote = "\"")


```



##Question 1. 
##Across the United States, which types of events (as indicated in the 'EVTYPE' variable) are most harmful with respect to population health?

###Conceptual basis
Population health has been defined as "the health outcomes of a group of individuals, including the distribution of such outcomes within the group".(1) 

 "Population health is a relatively new term that has not yet been precisely defined. Is it a concept of health or a field of study of health determinants?"(2)
 
Actually, there is no current universal standard to measure population health. Different researchers and organizations look at the measurement of population health from different points of views and use different tools and concepts to measure it; somehow, most of the concepts gravitating around Kindig and Stoddart definition of population health.

For the purpose of this project two measurements will be used 

**Fatalities:** as a mortality related measure. 
Based on the concept of "Years of potential life lost (YPLL)" 

Years of potential life lost (YPLL) or potential years of life lost (PYLL), is an estimate of the average years a person would have lived if he or she had not died prematurely.(3)

This report will interpret higher number of fatalities as more harmful with respect of population health.

**Injuries:** as a morbidity related measure. 
Based on the concepts of "Healthy Life Years (HLY)" and "Disability-free life expectancy (DFLE)"

Healthy Life Years  (HLY) measures the number of remaining years that a person of a certain age is expected to live without disability. It is actually a disability-free life expectancy.(4)

Disability-free life expectancy (DFLE)": The average number of years an individual is expected to live free of disability if current patterns of mortality and disability continue to apply. A statistical abstraction based on existing age-specific death rates and either age-specific disability prevalences or age-specific disability transition rates.(5)

This report will interpret higher numbers of injuries as an indicator of high occurrence of related disabilities and consequently, will interpret high numbers of injuries as more harmful with respect of population health.

***
(1) Wikipedia: http://en.wikipedia.org/wiki/Population_health -Retrieved 2014-09-20
(2) Kindig D, Stoddart G. What is population health? American Journal of Public Health 2003 Mar;93(3):380-3. http://www.ncbi.nlm.nih.gov/pubmed/12604476   Retrieved 2008-10-12.
(3) Gardner JW; Sanborn JS. "Years of potential life lost (YPLL)--what does it measure?". http://www.ncbi.nlm.nih.gov/pubmed/2083312 - Retrieved 2006-12-14.
(4) Wikipedia: http://en.wikipedia.org/wiki/Healthy_Life_Years - Retrieved 2014-09-20
(5) Glossary EurOhex website http://www.eurohex.eu/pdf/Glossary_v4.pdf - Retrieved 2014-09-20

***

###Program
In order to answer this question the following Sections are used:

####Section 1.1: 
####Creating a new sumarized data set 

```{r, echo =TRUE}

#    Section 1.1: Creating a new sumarized data set 
#         1. this code calculates the sum of fatalities and injuries 
#          by event type groups
#         2. add fatalities and injuries to ad a total of accidents
#         3. remove the groups with total accidents =0
#

library(plyr)
dataph<-ddply(data,.(EVTYPE),summarize,fatalities=sum(FATALITIES),injuries=sum(INJURIES))
dataph <-cbind(dataph,total=(dataph$fatalities+dataph$injuries))
dataph <- dataph[dataph$total > 0,]
```

####Section 1.2: 
####Ading a new column with standard event type names based in non standard historical event type names
Currently the table number 2.1.1 Storm Data Event Table (page 6) from the document NWSI 10-1615 of AUGUST 17, 2007 (data documentation),has a list of 48 standard Event Names.     

Nevertheless, in the historical data there are Event Names not contained in that current table 2.1.1.   

The code in this Section #2 adds a new column with the standard current event name as per the table 2.1.1 to the data set.

```{r, echo =TRUE}
#    Section 1.2: adding a new column with standard event type names 
#         based in non standard historical event type names


dataph[dataph$EVTYPE=="AVALANCE",5] <- "Avalanche"
dataph[dataph$EVTYPE=="AVALANCHE",5] <- "Avalanche"
dataph[dataph$EVTYPE=="BLACK ICE",5] <- "WinterWeather"
dataph[dataph$EVTYPE=="BLIZZARD",5] <- "Blizzard"
dataph[dataph$EVTYPE=="blowing snow",5] <- "HeavySnow"
dataph[dataph$EVTYPE=="BLOWING SNOW",5] <- "HeavySnow"
dataph[dataph$EVTYPE=="BRUSH FIRE",5] <- "Wildfire"
dataph[dataph$EVTYPE=="COASTAL FLOOD",5] <- "CoastalFlood"
dataph[dataph$EVTYPE=="Coastal Flooding",5] <- "CoastalFlood"
dataph[dataph$EVTYPE=="COASTAL FLOODING",5] <- "CoastalFlood"
dataph[dataph$EVTYPE=="COASTAL FLOODING/EROSION",5] <- "CoastalFlood"
dataph[dataph$EVTYPE=="Coastal Storm",5] <- "StormSurge/Tide"
dataph[dataph$EVTYPE=="COASTAL STORM",5] <- "StormSurge/Tide"
dataph[dataph$EVTYPE=="COASTALSTORM",5] <- "StormSurge/Tide"
dataph[dataph$EVTYPE=="Cold",5] <- "Cold/Windhill"
dataph[dataph$EVTYPE=="COLD",5] <- "Cold/Windhill"
dataph[dataph$EVTYPE=="COLD AND SNOW",5] <- "WinterWeather"
dataph[dataph$EVTYPE=="Cold Temperature",5] <- "Cold/Windchill"
dataph[dataph$EVTYPE=="COLD WAVE",5] <- "Cold/Windchill"
dataph[dataph$EVTYPE=="COLD WEATHER",5] <- "WinterWeather"
dataph[dataph$EVTYPE=="COLD/WIND CHILL",5] <- "Cold/Windchill"
dataph[dataph$EVTYPE=="COLD/WINDS",5] <- "Cold/Windhill"
dataph[dataph$EVTYPE=="DENSE FOG",5] <- "DenseFog"
dataph[dataph$EVTYPE=="DROUGHT",5] <- "Excessive Heat"
dataph[dataph$EVTYPE=="DROUGHT/EXCESSIVE HEAT",5] <- "Excessive Heat"
dataph[dataph$EVTYPE=="DROWNING",5] <- "N/A"
dataph[dataph$EVTYPE=="DRY MICROBURST",5] <- "DustStorm"
dataph[dataph$EVTYPE=="DRY MIRCOBURST WINDS",5] <- "DustStorm"
dataph[dataph$EVTYPE=="Dust Devil",5] <- "DustDevil"
dataph[dataph$EVTYPE=="DUST DEVIL",5] <- "DustDevil"
dataph[dataph$EVTYPE=="DUST STORM",5] <- "DustStorm"
dataph[dataph$EVTYPE=="EXCESSIVE HEAT",5] <- "Excessive Heat"
dataph[dataph$EVTYPE=="EXCESSIVE RAINFALL",5] <- "HeavyRain"
dataph[dataph$EVTYPE=="EXCESSIVE SNOW",5] <- "HeavySnow"
dataph[dataph$EVTYPE=="Extended Cold",5] <- "Extreme Cold / Windchill"
dataph[dataph$EVTYPE=="Extreme Cold",5] <- "Extreme Cold / Windchill"
dataph[dataph$EVTYPE=="EXTREME COLD",5] <- "Extreme Cold / Windchill"
dataph[dataph$EVTYPE=="EXTREME COLD/WIND CHILL",5] <- "Extreme Cold / Windchill"
dataph[dataph$EVTYPE=="EXTREME HEAT",5] <- "Excessive Heat"
dataph[dataph$EVTYPE=="EXTREME WINDCHILL",5] <- "Extremeold/Windhill"
dataph[dataph$EVTYPE=="FALLING SNOW/ICE",5] <- "HeavySnow"
dataph[dataph$EVTYPE=="FLASH FLOOD",5] <- "Flash Flood"
dataph[dataph$EVTYPE=="FLASH FLOOD/FLOOD",5] <- "Flash Flood"
dataph[dataph$EVTYPE=="FLASH FLOODING",5] <- "Flash Flood"
dataph[dataph$EVTYPE=="FLASH FLOODING/FLOOD",5] <- "Flash Flood"
dataph[dataph$EVTYPE=="FLASH FLOODS",5] <- "Flash Flood"
dataph[dataph$EVTYPE=="FLOOD",5] <- "Flood"
dataph[dataph$EVTYPE=="FLOOD & HEAVY RAIN",5] <- "Flood"
dataph[dataph$EVTYPE=="FLOOD/FLASH FLOOD",5] <- "Flood"
dataph[dataph$EVTYPE=="FLOOD/RIVER FLOOD",5] <- "Flood"
dataph[dataph$EVTYPE=="FLOODING",5] <- "Flood"
dataph[dataph$EVTYPE=="FOG",5] <- "DenseFog"
dataph[dataph$EVTYPE=="FOG AND COLD TEMPERATURES",5] <- "DenseFog"
dataph[dataph$EVTYPE=="FREEZE",5] <- "Frost/Freeze"
dataph[dataph$EVTYPE=="FREEZING DRIZZLE",5] <- "Frost/Freeze"
dataph[dataph$EVTYPE=="FREEZING RAIN",5] <- "Frost/Freeze"
dataph[dataph$EVTYPE=="FREEZING RAIN/SNOW",5] <- "Frost/Freeze"
dataph[dataph$EVTYPE=="Freezing Spray",5] <- "Frost/Freeze"
dataph[dataph$EVTYPE=="FROST",5] <- "Frost/Freeze"
dataph[dataph$EVTYPE=="FUNNEL CLOUD",5] <- "Funnelloud"
dataph[dataph$EVTYPE=="GLAZE",5] <- "Ice Storm"
dataph[dataph$EVTYPE=="GLAZE/ICE STORM",5] <- "Ice Storm"
dataph[dataph$EVTYPE=="GUSTY WIND",5] <- "StrongWind"
dataph[dataph$EVTYPE=="Gusty winds",5] <- "StrongWind"
dataph[dataph$EVTYPE=="Gusty Winds",5] <- "StrongWind"
dataph[dataph$EVTYPE=="GUSTY WINDS",5] <- "StrongWind"
dataph[dataph$EVTYPE=="HAIL",5] <- "Hail"
dataph[dataph$EVTYPE=="HAZARDOUS SURF",5] <- "HighSurf"
dataph[dataph$EVTYPE=="HEAT",5] <- "Heat"
dataph[dataph$EVTYPE=="Heat Wave",5] <- "Heat"
dataph[dataph$EVTYPE=="HEAT WAVE",5] <- "Heat"
dataph[dataph$EVTYPE=="HEAT WAVE DROUGHT",5] <- "Heat"
dataph[dataph$EVTYPE=="HEAT WAVES",5] <- "Heat"
dataph[dataph$EVTYPE=="HEAVY RAIN",5] <- "HeavyRain"
dataph[dataph$EVTYPE=="HEAVY RAINS",5] <- "HeavyRain"
dataph[dataph$EVTYPE=="HEAVY SEAS",5] <- "HighSurf"
dataph[dataph$EVTYPE=="HEAVY SNOW",5] <- "HeavySnow"
dataph[dataph$EVTYPE=="HEAVY SNOW AND HIGH WINDS",5] <- "HeavySnow"
dataph[dataph$EVTYPE=="Heavy snow shower",5] <- "HeavySnow"
dataph[dataph$EVTYPE=="HEAVY SNOW/BLIZZARD/AVALANCHE",5] <- "HeavySnow"
dataph[dataph$EVTYPE=="HEAVY SNOW/ICE",5] <- "HeavySnow"
dataph[dataph$EVTYPE=="Heavy Surf",5] <- "HighSurf"
dataph[dataph$EVTYPE=="HEAVY SURF",5] <- "HighSurf"
dataph[dataph$EVTYPE=="Heavy surf and wind",5] <- "HighSurf"
dataph[dataph$EVTYPE=="HEAVY SURF/HIGH SURF",5] <- "HighSurf"
dataph[dataph$EVTYPE=="HIGH",5] <- "HighSurf"
dataph[dataph$EVTYPE=="HIGH SEAS",5] <- "HighSurf"
dataph[dataph$EVTYPE=="High Surf",5] <- "HighSurf"
dataph[dataph$EVTYPE=="HIGH SURF",5] <- "HighSurf"
dataph[dataph$EVTYPE=="HIGH SWELLS",5] <- "HighSurf"
dataph[dataph$EVTYPE=="HIGH WATER",5] <- "HighSurf"
dataph[dataph$EVTYPE=="HIGH WAVES",5] <- "HighSurf"
dataph[dataph$EVTYPE=="HIGH WIND",5] <- "High Wind"
dataph[dataph$EVTYPE=="HIGH WIND 48",5] <- "High Wind"
dataph[dataph$EVTYPE=="HIGH WIND AND SEAS",5] <- "High Wind"
dataph[dataph$EVTYPE=="HIGH WIND/HEAVY SNOW",5] <- "High Wind"
dataph[dataph$EVTYPE=="HIGH WIND/SEAS",5] <- "High Wind"
dataph[dataph$EVTYPE=="HIGH WINDS",5] <- "High Wind"
dataph[dataph$EVTYPE=="HIGH WINDS/COLD",5] <- "High Wind"
dataph[dataph$EVTYPE=="HIGH WINDS/SNOW",5] <- "High Wind"
dataph[dataph$EVTYPE=="HURRICANE",5] <- "Hurricane(Typhoon)"
dataph[dataph$EVTYPE=="HURRICANE-GENERATED SWELLS",5] <- "Hurricane(Typhoon)"
dataph[dataph$EVTYPE=="Hurricane Edouard",5] <- "Hurricane(Typhoon)"
dataph[dataph$EVTYPE=="HURRICANE EMILY",5] <- "Hurricane(Typhoon)"
dataph[dataph$EVTYPE=="HURRICANE ERIN",5] <- "Hurricane(Typhoon)"
dataph[dataph$EVTYPE=="HURRICANE FELIX",5] <- "Hurricane(Typhoon)"
dataph[dataph$EVTYPE=="HURRICANE OPAL",5] <- "Hurricane(Typhoon)"
dataph[dataph$EVTYPE=="HURRICANE OPAL/HIGH WINDS",5] <- "Hurricane(Typhoon)"
dataph[dataph$EVTYPE=="HURRICANE/TYPHOON",5] <- "Hurricane(Typhoon)"
dataph[dataph$EVTYPE=="HYPERTHERMIA/EXPOSURE",5] <- "Excessive Heat"
dataph[dataph$EVTYPE=="HYPOTHERMIA",5] <- "Extremeold/Windhill"
dataph[dataph$EVTYPE=="Hypothermia/Exposure",5] <- "Extremeold/Windhill"
dataph[dataph$EVTYPE=="HYPOTHERMIA/EXPOSURE",5] <- "Extremeold/Windhill"
dataph[dataph$EVTYPE=="ICE",5] <- "WinterWeather"
dataph[dataph$EVTYPE=="ICE ON ROAD",5] <- "WinterWeather"
dataph[dataph$EVTYPE=="ICE ROADS",5] <- "WinterWeather"
dataph[dataph$EVTYPE=="ICE STORM",5] <- "Ice Storm"
dataph[dataph$EVTYPE=="ICE STORM/FLASH FLOOD",5] <- "Ice Storm"
dataph[dataph$EVTYPE=="ICY ROADS",5] <- "WinterWeather"
dataph[dataph$EVTYPE=="LANDSLIDE",5] <- "Flood"
dataph[dataph$EVTYPE=="LANDSLIDES",5] <- "Flood"
dataph[dataph$EVTYPE=="LIGHT SNOW",5] <- "WinterWeather"
dataph[dataph$EVTYPE=="LIGHTNING",5] <- "Lightning"
dataph[dataph$EVTYPE=="LIGHTNING AND THUNDERSTORM WIN",5] <- "Lightning"
dataph[dataph$EVTYPE=="LIGHTNING INJURY",5] <- "Lightning"
dataph[dataph$EVTYPE=="LIGHTNING.",5] <- "Lightning"
dataph[dataph$EVTYPE=="LOW TEMPERATURE",5] <- "Cold/Windhill"
dataph[dataph$EVTYPE=="Marine Accident",5] <- "N/A"
dataph[dataph$EVTYPE=="MARINE HIGH WIND",5] <- "MarineHighWind"
dataph[dataph$EVTYPE=="MARINE MISHAP",5] <- "N/A"
dataph[dataph$EVTYPE=="MARINE STRONG WIND",5] <- "MarineStrongWind"
dataph[dataph$EVTYPE=="MARINE THUNDERSTORM WIND",5] <- "MarineThunderstormWind"
dataph[dataph$EVTYPE=="MARINE TSTM WIND",5] <- "MarineThunderstormWind"
dataph[dataph$EVTYPE=="MINOR FLOODING",5] <- "Flood"
dataph[dataph$EVTYPE=="MIXED PRECIP",5] <- "HeavyRain"
dataph[dataph$EVTYPE=="Mudslide",5] <- "Flood"
dataph[dataph$EVTYPE=="Mudslides",5] <- "Flood"
dataph[dataph$EVTYPE=="NON-SEVERE WIND DAMAGE",5] <- "High Wind"
dataph[dataph$EVTYPE=="NON TSTM WIND",5] <- "High Wind"
dataph[dataph$EVTYPE=="OTHER",5] <- "N/A"
dataph[dataph$EVTYPE=="RAIN/SNOW",5] <- "HeavyRain"
dataph[dataph$EVTYPE=="RAIN/WIND",5] <- "HeavyRain"
dataph[dataph$EVTYPE=="RAPIDLY RISING WATER",5] <- "CoastalFlood"
dataph[dataph$EVTYPE=="RECORD COLD",5] <- "Extremeold/Windhill"
dataph[dataph$EVTYPE=="RECORD HEAT",5] <- "Excessive Heat"
dataph[dataph$EVTYPE=="RECORD/EXCESSIVE HEAT",5] <- "Excessive Heat"
dataph[dataph$EVTYPE=="RIP CURRENT",5] <- "Rip Current"
dataph[dataph$EVTYPE=="RIP CURRENTS",5] <- "Rip Current"
dataph[dataph$EVTYPE=="RIP CURRENTS/HEAVY SURF",5] <- "Rip Current"
dataph[dataph$EVTYPE=="RIVER FLOOD",5] <- "Flood"
dataph[dataph$EVTYPE=="River Flooding",5] <- "Flood"
dataph[dataph$EVTYPE=="RIVER FLOODING",5] <- "Flood"
dataph[dataph$EVTYPE=="ROGUE WAVE",5] <- "HighSurf"
dataph[dataph$EVTYPE=="ROUGH SEAS",5] <- "HighSurf"
dataph[dataph$EVTYPE=="ROUGH SURF",5] <- "HighSurf"
dataph[dataph$EVTYPE=="SLEET",5] <- "Ice Storm"
dataph[dataph$EVTYPE=="SMALL HAIL",5] <- "Hail"
dataph[dataph$EVTYPE=="Snow",5] <- "HeavySnow"
dataph[dataph$EVTYPE=="SNOW",5] <- "HeavySnow"
dataph[dataph$EVTYPE=="SNOW AND ICE",5] <- "HeavySnow"
dataph[dataph$EVTYPE=="SNOW SQUALL",5] <- "HeavySnow"
dataph[dataph$EVTYPE=="Snow Squalls",5] <- "HeavySnow"
dataph[dataph$EVTYPE=="SNOW/ BITTER COLD",5] <- "HeavySnow"
dataph[dataph$EVTYPE=="SNOW/HIGH WINDS",5] <- "HeavySnow"
dataph[dataph$EVTYPE=="STORM SURGE",5] <- "StormSurge/Tide"
dataph[dataph$EVTYPE=="STORM SURGE/TIDE",5] <- "StormSurge/Tide"
dataph[dataph$EVTYPE=="STRONG WIND",5] <- "High Wind"
dataph[dataph$EVTYPE=="Strong Winds",5] <- "High Wind"
dataph[dataph$EVTYPE=="STRONG WINDS",5] <- "High Wind"
dataph[dataph$EVTYPE=="THUNDERSNOW",5] <- "Thunderstorm Wind"
dataph[dataph$EVTYPE=="THUNDERSTORM",5] <- "Thunderstorm Wind"
dataph[dataph$EVTYPE=="THUNDERSTORM  WINDS",5] <- "Thunderstorm Wind"
dataph[dataph$EVTYPE=="THUNDERSTORM WIND",5] <- "Thunderstorm Wind"
dataph[dataph$EVTYPE=="THUNDERSTORM WIND (G40)",5] <- "Thunderstorm Wind"
dataph[dataph$EVTYPE=="THUNDERSTORM WIND G52",5] <- "Thunderstorm Wind"
dataph[dataph$EVTYPE=="THUNDERSTORM WINDS",5] <- "Thunderstorm Wind"
dataph[dataph$EVTYPE=="THUNDERSTORM WINDS 13",5] <- "Thunderstorm Wind"
dataph[dataph$EVTYPE=="THUNDERSTORM WINDS/HAIL",5] <- "Thunderstorm Wind"
dataph[dataph$EVTYPE=="THUNDERSTORM WINDSS",5] <- "Thunderstorm Wind"
dataph[dataph$EVTYPE=="THUNDERSTORMS WINDS",5] <- "Thunderstorm Wind"
dataph[dataph$EVTYPE=="THUNDERSTORMW",5] <- "Thunderstorm Wind"
dataph[dataph$EVTYPE=="THUNDERTORM WINDS",5] <- "Thunderstorm Wind"
dataph[dataph$EVTYPE=="TIDAL FLOODING",5] <- "Flood"
dataph[dataph$EVTYPE=="TORNADO",5] <- "Tornado"
dataph[dataph$EVTYPE=="TORNADO F2",5] <- "Tornado"
dataph[dataph$EVTYPE=="TORNADO F3",5] <- "Tornado"
dataph[dataph$EVTYPE=="TORNADOES, TSTM WIND, HAIL",5] <- "Tornado"
dataph[dataph$EVTYPE=="Torrential Rainfall",5] <- "HeavyRain"
dataph[dataph$EVTYPE=="TROPICAL STORM",5] <- "Thunderstorm Wind"
dataph[dataph$EVTYPE=="TROPICAL STORM GORDON",5] <- "Thunderstorm Wind"
dataph[dataph$EVTYPE=="TSTM WIND",5] <- "Thunderstorm Wind"
dataph[dataph$EVTYPE=="TSTM WIND (G35)",5] <- "Thunderstorm Wind"
dataph[dataph$EVTYPE=="TSTM WIND (G40)",5] <- "Thunderstorm Wind"
dataph[dataph$EVTYPE=="TSTM WIND (G45)",5] <- "Thunderstorm Wind"
dataph[dataph$EVTYPE=="TSTM WIND/HAIL",5] <- "Thunderstorm Wind"
dataph[dataph$EVTYPE=="TSUNAMI",5] <- "Tsunami"
dataph[dataph$EVTYPE=="TYPHOON",5] <- "Hurricane(Typhoon)"
dataph[dataph$EVTYPE=="UNSEASONABLY COLD",5] <- "Cold/Windhill"
dataph[dataph$EVTYPE=="UNSEASONABLY WARM",5] <- "Heat"
dataph[dataph$EVTYPE=="UNSEASONABLY WARM AND DRY",5] <- "Heat"
dataph[dataph$EVTYPE=="URBAN AND SMALL STREAM FLOODIN",5] <- "Flood"
dataph[dataph$EVTYPE=="URBAN/SML STREAM FLD",5] <- "Flood"
dataph[dataph$EVTYPE=="WARM WEATHER",5] <- "Heat"
dataph[dataph$EVTYPE=="WATERSPOUT",5] <- "Tornado"
dataph[dataph$EVTYPE=="WATERSPOUT TORNADO",5] <- "Tornado"
dataph[dataph$EVTYPE=="WATERSPOUT/TORNADO",5] <- "Tornado"
dataph[dataph$EVTYPE=="Whirlwind",5] <- "Tornado"
dataph[dataph$EVTYPE=="WILD FIRES",5] <- "Wildfire"
dataph[dataph$EVTYPE=="WILD/FOREST FIRE",5] <- "Wildfire"
dataph[dataph$EVTYPE=="WILDFIRE",5] <- "Wildfire"
dataph[dataph$EVTYPE=="WIND",5] <- "High Wind"
dataph[dataph$EVTYPE=="WIND STORM",5] <- "StrongWind"
dataph[dataph$EVTYPE=="WINDS",5] <- "StrongWind"
dataph[dataph$EVTYPE=="WINTER STORM",5] <- "WinterStorm"
dataph[dataph$EVTYPE=="WINTER STORM HIGH WINDS",5] <- "WinterStorm"
dataph[dataph$EVTYPE=="WINTER STORMS",5] <- "WinterStorm"
dataph[dataph$EVTYPE=="WINTER WEATHER",5] <- "WinterWeather"
dataph[dataph$EVTYPE=="WINTER WEATHER MIX",5] <- "WinterWeather"
dataph[dataph$EVTYPE=="WINTER WEATHER/MIX",5] <- "WinterWeather"
dataph[dataph$EVTYPE=="WINTRY MIX",5] <- "WinterWeather"

colnames(dataph)[5] <- "stdevtype"
```


####Section 1.3: 
####Summarizing/aggregating number of fatalities and injuries by event type


```{r, echo =TRUE}
#    Section 1.3: summarizing/aggregating number of fatalities and injuries 
#         by the new standardized event type

dataph<-ddply(dataph,.(stdevtype),summarize,fatalities=sum(fatalities),injuries=sum(injuries))
```


####Section 1.4: 
####Data set format change for plotting graph 
There is no major transformation in the values of the data in this step. 
The transformation occur in the format of the data. That is in order to meet the requirements of the plot graphics functions.



```{r, echo =TRUE}
#
#    Section 1.4: Data set format change for plotting graph 
#         1. The data is separated in two matrices (dataf and fatai) 
#              one for Fatalities data and other for Injuries data.
#         2. labels are added to the new matices.
#         3. Values are divided by 1000 for formating reasons.
#         4. rows are ordered by value.
#         5. data set is trasposed
#         6. the dataset is restricted to the top 10 higher values.


dataf<-matrix(dataph$fatalities/1000)
dimnames(dataf) <- list(dataph$stdevtype, "fatalities")
dataf <- dataf[order(dataf[,1]),] 
dataf<-t(dataf)
dataf <- dataf[,c((length(dataf[1,])-10):(length(dataf[1,])))]

datai<-matrix(dataph$injuries/1000)
dimnames(datai) <- list(dataph$stdevtype, "injuries")
datai <- datai[order(datai[,1]),] 
datai<-t(datai)
datai <- datai[,c((length(datai[1,])-10):(length(datai[1,])))]

```

####Section 1.5: 
####Plotting graphs


```{r, echo =TRUE}
#    Section 1.5: Plotting graphs
{
par(las=1) # make label text perpendicular to axis
par(mar=c(5,8,4,2)) # increase y-axis margin.
par(ps = 10, cex = 1,cex.main = 1.5, cex.axis = 1, cex.lab = 1)

barplot(dataf, main="Graph #1: Total weather related fatalities in the US\n by the 10 most harmful weather event types\n registered betwen 1950 and 2011",
         xlab="'000s Fatalities", col=c("grey"),
         ,horiz=TRUE,xlim=c(0,6))

barplot(datai, main="Graph #2: Total weather related injuries in the US\n by the 10 most harmful weather event types\n registered betwen 1950 and 2011",
         xlab="'000s Injuries", col=c("grey"),
         ,horiz=TRUE,xlim=c(0,100))
}
```

####Section 1.6: Preparing results for printing
In this step the list of results get ordered.

```{r, echo =TRUE}
#    Section 1.6: Preparing results for printing

tdataf <- t(dataf)
tdataf <- t(tdataf)
fevents <- labels(tdataf[order(-tdataf[,1]),]) 

tdatai <- t(datai)
tdatai <- t(tdatai)
ievents <- labels(tdatai[order(-tdatai[,1]),]) 
```

####Section 1.7: RESULTS (Printing results)

```{r, echo =TRUE}

#    Section 1.7: Printing Results
title <- paste ("***  RESULTS  ***")
question1 <- paste("Question #1. Across the United States, which types of 
                   events (as indicated in the 'EVTYPE' variable) are most harmful
                   with respect to population health?")
result1 <- paste("As shown in the Graph #1, in terms of Fatalities, the 10 most
                 harmful weather event types across the US between 1950 and 2011 
                 are: ",paste(fevents,collapse=", "))
result2 <- paste("As shown in the Graph #2, In terms of Injuries, the 10 most harmful
                 weather event types across the US between 1950 and 2011 
                 are: ",paste(ievents,collapse=", "))
{nicecomment(title)
nicecomment(question1)
nicecomment(result1)
nicecomment(result2)}

```


##Question 2. 
##Across the United States, which types of events have the greatest economic consequences?

###Conceptual basis
For this question the fields PROPDMG (Property damages in US\$) and CROPDMG (Crop Damages in US\$) are added on every row of the original data to calculate the Total Economic Damage (in US$) by event. Then the Total Economic Damages are aggregated by event type.


This report will interpret the event types with higher calculated Total Economic Damages as having greater economic consequences.

###Program
In order to answer this question the following Sections are used:

####Section 2.1: 
####Creating a new sumarized data set 

```{r, echo =TRUE}
#    Section 2.1: Creating a new sumarized data set 
#         1. this code calculates the sum of fatalities and injuries 
#              on event type groups
#         2. add fatalities and injuries to ad a total of accidents
#         3. remove the groups with total accidents =0


data[data$PROPDMGEXP=="",38] <- 1
data[data$PROPDMGEXP=="K",38] <- 1000
data[data$PROPDMGEXP=="M",38] <- 1000000
data[data$PROPDMGEXP=="B",38] <- 1000000000
data[is.na(data$CROPDMGEXP),39] <- 1
data[data$CROPDMGEXP=="K" & !is.na(data$CROPDMGEXP),39] <- 1000
data[data$CROPDMGEXP=="M" & !is.na(data$CROPDMGEXP),39] <- 1000000
data[data$CROPDMGEXP=="B" & !is.na(data$CROPDMGEXP),39] <- 1000000000
data[,40]<- (data[,25] * data[,38]) + (data[,27] * data[,39])

datae <- ddply(data,.(EVTYPE),summarize,econdam=sum(V40))
datae <- datae[datae$econdam > 0 & !is.na(datae$econdam),]

```

####Section 2.2: 
####Adding a new column with standard event type names based in non standard historical event type names
Currently the table number 2.1.1 Storm Data Event Table (page 6) from the document NWSI 10-1615 of AUGUST 17, 2007 (data documentation),has a list of 48 standard Event Names.    

Nevertheless, in the historical data there are Event Names not contained in that current table 2.1.1.   

The code in this Section #2 adds a new column with the standard current event name as per the table 2.1.1 to the data set.

```{r, echo =TRUE}
#    Section 2.2: adding a new column with standard event type names 
#         based in non standard historical event type names

datae[datae$EVTYPE=="ASTRONOMICAL LOW TIDE",3] <- "Astronomical Low Tide"
datae[datae$EVTYPE=="COLD AIR TORNADO",3] <- "Tornado"
datae[datae$EVTYPE=="COLD AND WET CONDITIONS",3] <- "Winter Weather"
datae[datae$EVTYPE=="COOL AND WET",3] <- "Winter Weather"
datae[datae$EVTYPE=="Damaging Freeze",3] <- "Frost/Freeze"
datae[datae$EVTYPE=="DUST STORM/HIGH WINDS",3] <- "Dust Storm"
datae[datae$EVTYPE=="Early Frost",3] <- "Frost/Freeze"
datae[datae$EVTYPE=="EXCESSIVE WETNESS",3] <- "Heavy Rain"
datae[datae$EVTYPE=="FLASH FLOODING/FLOOD",3] <- "Flash Flood"
datae[datae$EVTYPE=="FOREST FIRES",3] <- "Wildfire"
datae[datae$EVTYPE=="Freeze",3] <- "Frost/Freeze"
datae[datae$EVTYPE=="Frost/Freeze",3] <- "Frost/Freeze"
datae[datae$EVTYPE=="GLAZE ICE",3] <- "Winter Weather"
datae[datae$EVTYPE=="HAIL 075",3] <- "Hail"
datae[datae$EVTYPE=="HAIL 125",3] <- "Hail"
datae[datae$EVTYPE=="HAIL 200",3] <- "Hail"
datae[datae$EVTYPE=="HAIL/WINDS",3] <- "Hail"
datae[datae$EVTYPE=="HEAT WAVE DROUGHT",3] <- "Excessive Heat"
datae[datae$EVTYPE=="Heavy Rain/High Surf",3] <- "Heavy Rain"
datae[datae$EVTYPE=="HEAVY SNOW/HIGH WINDS & FLOOD",3] <- "Heavy Snow"
datae[datae$EVTYPE=="HIGH WINDS HEAVY RAINS",3] <- "High Wind"
datae[datae$EVTYPE=="HIGH WINDS/COLD",3] <- "High Wind"
datae[datae$EVTYPE=="HURRICANE OPAL/HIGH WINDS",3] <- "Hurricane (Typhoon)"
datae[datae$EVTYPE=="LAKESHORE FLOOD",3] <- "Flood"
datae[datae$EVTYPE=="MARINE THUNDERSTORM WIND",3] <- "Marine Thunderstorm Wind"
datae[datae$EVTYPE=="THUNDERSTORM HAIL",3] <- "Thunderstorm Wind"
datae[datae$EVTYPE=="THUNDERSTORM WIND.",3] <- "Thunderstorm Wind"
datae[datae$EVTYPE=="THUNDERSTORM WINDS G60",3] <- "Thunderstorm Wind"
datae[datae$EVTYPE=="TORNADOES",3] <- "Tornado"
datae[datae$EVTYPE=="TORNADOES, TSTM WIND, HAIL",3] <- "Tornado"
datae[datae$EVTYPE=="TROPICAL STORM GORDON",3] <- "Tropical Storm"
datae[datae$EVTYPE=="Unseasonable Cold",3] <- "Cold/Wind Chill"
datae[datae$EVTYPE=="UNSEASONAL RAIN",3] <- "Heavy Rain"
datae[datae$EVTYPE=="WILD/FOREST FIRES",3] <- "Wildfire"
datae[datae$EVTYPE=="WINTER STORM HIGH WINDS",3] <- "Winter Storm"

colnames(datae)[3] <- "stdevtype"
```

####Section 2.3: 
####Summarizing/aggregating total economic damages by event type

```{r, echo =TRUE}
#    Section 2.3: Summarizing/aggregating total economic damages by event type

datae<-ddply(datae,.(stdevtype),summarize,econdam=sum(econdam))
```

####Section 2.4: 
####Data set format change for plotting graph 
There is no major transformation in the values of the data in this step. 
The transformation occur in the format of the data. That is in order to meet the requirements of the plot graphics functions.



```{r, echo =TRUE}
#
#    Section 2.4: Data set format change for plotting graph 
#         1. The data is separated into a new matrix (datae) 
#         2. labels are added to the new matix.
#         3. Values are divided by 1,000,000 for formating reasons.
#         4. rows are ordered by value.
#         5. data set is trasposed
#         6. the dataset is restricted to the top 10 higher values.

dataed <- matrix(datae$econdam/1000000)
dimnames(dataed) <- list(datae$stdevtype, "EconomicalDamage")
dataed <- dataed[order(dataed[,1]),] 
dataed<-t(dataed)
dataed <- dataed[,c((length(dataed[1,])-10):(length(dataed[1,])))]
```

####Section 2.5: 
####Plotting graph


```{r, echo =TRUE}
#    Section 2.5: Plotting graph

par(las=1) # make label text perpendicular to axis
par(mar=c(5,10,4,2)) # increase y-axis margin.
par(ps = 10, cex = 1,cex.main = 1.5, cex.axis = 1, cex.lab = 1)
barplot(dataed, main="Graph #3: Total weather related economical damage across the US\n by the 10 weather event types with greatest economic consecuences\nregistered betwen 1950 and 2011",
         xlab="'000,000s US$", col=c("grey"),
         ,horiz=TRUE,xlim=c(0,2000))
```


####Section 2.6: 
####Preparing results for printing
In this step the list of results get ordered.

```{r, echo =TRUE}
#    Section 2.6: Preparing results for printing 
tdataed <- t(dataed)
tdataed <- t(tdataed)
eevents <- labels(tdataed[order(-tdataed[,1]),])
```

####Section 2.7: RESULTS (Printing results)

```{r, echo =TRUE}
#    Section 2.7: Printing Results
question1 <- paste("Question #2. Across the United States, which types of events have the greatest economic consequences?")
result1 <- paste("As shown in the Graph #3, in terms of economical damage to property and crops, the 10 most harmful weather event types across the US between 1950 and 2011 are: ",paste(eevents,collapse=", "))

{nicecomment(title)
nicecomment(question1)
nicecomment(result1)}
```


 